From Archives to AI: Residential Property Data Across Three Decades in Brunei Darussalam

### 2025-03-05 manuscript.qmd
Final tweaks manuscript

---
title: 'From Archives to AI: Residential Property Data Across Three Decades in Brunei Darussalam'
authors:
  - name: Haziq Jamil
    orcid: 0000-0003-3298-1010
    email: haziq.jamil@ubd.edu.bn
    url: 'https://haziqj.ml'
    corresponding: true
    affiliations:
      - name: Universiti Brunei Darussalam
        id: ubd
        department: Mathematical Sciences, Faculty of Science
        address: Universiti Brunei Darussalam, Jalan Tungku Link
        city: Bandar Seri Begawan
        country: Brunei
        postal-code: BE 1410
  - name: Amira Barizah Noorosmawie
    affiliations:
      - ref: ubd        
  - name: Hafeezul Waezz Rabu
    affiliations:
      - ref: ubd
  - name: Lutfi Abdul Razak
    orcid: 0000-0003-1681-0895
    email: lutfi.razak@ubd.edu.bn
    affiliations:
      - name: UBD School of Business and Economics
        id: ubdsbe
        # department: Mathematical Sciences, Faculty of Science
        address: Universiti Brunei Darussalam, Jalan Tungku Link
        city: Bandar Seri Begawan
        country: Brunei
        postal-code: BE 1410         
date-modified: last-modified
abstract: |
  This article introduces the first publicly available data set for analyzing the Brunei housing market, covering more than 30,000 property listings from 1993 to early 2025. The data set, curated from property advertisements in newspapers and online platforms, includes key attributes such as price, location, property type, and physical characteristics, enriched with area-level spatial information. Comprehensive and historical, it complements the Brunei Darussalam Central Bank’s Residential Property Price Index (RPPI), addressing the limitations of restricted access to raw RPPI data and its relatively short timeline since its inception in 2015. Data collection involved manual transcription from archival sources and automated web scraping using programmatic methods, supported by innovative cleaning with Large Language Models (LLMs) to structure unformatted text. The data set enables spatial and temporal analysis, with potential applications in economics, urban planning, and real estate research. Although listing prices serve only as a proxy for market values and may deviate from actual sale prices due to negotiation dynamics and other factors, this data set still provides a valuable resource for reproducible analysis of housing market trends and for informing policy decisions.
keywords:
  - Housing Market
  - Brunei
  - Property Listings
  - Spatial Data
  - Web Scraping
  - Large Language Models
---

@@

- **First of its kind for Brunei**. This data set is the first publicly available compilation of Brunei’s residential property data, covering `{r} scales::comma(nrow(hsp))` listings from `r min(years)` to early `r max(years)`. Previous studies on Brunei's housing have been limited to recent or non-spatial data [@ng2022characterising;@jamil2025leveraging], or qualitative in nature due to the lack of structured data [@hassan2011cultural;@hassan2023sociocultural]. Researchers may use this data to explore temporal variations in property listings and to compare housing characteristics across different regions within Brunei.

<!-- - **First of its kind for Brunei**. This data set enables comprehensive spatial and temporal analysis of the Brunei housing market. To our knowledge, no other publicly available data set exists for this purpose, making it a significant contribution to housing market research. Previous studies on Brunei's housing have been limited to recent or non-spatial data [@shabunko2014developing; @ng2022characterising; @jamil2025leveraging], or qualitative in nature due to the lack of a structured data set [@hassan2011cultural; @hassan2023sociocultural]. This aligns with the growing trend of using data analytics in the real estate industry [@braesemann2020proptech; @delisle2020big]. -->

- **Historical and spatial insights**. Spanning over three decades, this data set allows for the study of long-term housing trends in Brunei. The spatial information available enables analysis at various administrative levels, providing insights into urban development patterns, regional differences, and the evolution of housing characteristics over time. This historical depth is particularly valuable given the lack of previous data before the establishment of the Residential Property Price Index (RPPI) [@bdcb2021technical] in 2015.

@@

- **Opportunities for handling of missing data**. The data set includes instances of missing or inconsistent data, particularly in house characteristics, which present opportunities for further research on data imputation techniques. Researchers may explore advanced statistical and machine learning methods to address missing values and improve data reliability. The structured format and detailed documentation of the data set support such methodological investigations, making it a valuable testbed for developing and evaluating new approaches to handling incomplete data in real-world data sets.

<!-- - **Application of Large Language Models (LLMs)**. This project highlights an innovative use of Large Language Models (LLMs) for data cleaning. The LLM was employed to extract structured information from unstructured text descriptions with great accuracy. This method significantly reduced the time and effort required for data cleaning. It can also be applied to other data sets with unstructured text, such as social media or document archives, to extract valuable information for analysis. -->

@@

Finally, metadata about the property is included for transparency and informational purposes. 
The variable agent specifies the (anonymised) identifier of the real estate agent or agency responsible for the listing, while source identifies the platform or medium from which the listing was obtained, such as a newspaper, magazine, or website. 
The method variable details the data collection approach, which is further elaborated in the section below.

@@

The date variable in the data set refers to the date on which the property listing was obtained. 
It is important to note that this is not the date the property was sold, nor does it necessarily reflect the precise timing of other transactions related to the property. 
Users should set their expectations accordingly, as the primary purpose of the date is to capture the state of the housing market at a specific point in time.

For analysis, we recommend aggregating data by quarters, as represented by the quarter variable. 
This aggregation helps address potential issues like missing data (see subsection below) and provides a more stable and robust representation of market trends, making it suitable for temporal analysis of the housing market.

@@

This data set endeavours to provide complete information regarding the listing date, spatial information, advertised price, and metadata. 
Nonetheless, we hereby report that missing values are present across various **property characteristics**, including type, tenure, status, plot area, floor area, storeys, beds, and baths. 
The reason for this is due to the nature of the property advertisements, which may not always include complete information when advertised by the real estate agents. 
Missing values are represented by blank cells in the CSV file, and the severity of missing values is summarised in @tbl-avail. 
In summary, 10.5% of the records contain missing values for all key house characteristics (i.e. plot area, floor area, beds, and baths), which, depending on the research question, may necessitate imputation or exclusion of these records.

# Experimental Design, Materials and Methods

@@

- `price` variables were cleaned by removing non-numeric characters and converted to integers.
- `beds` and `baths` were converted to integers.
- `date` variables were formatted properly as `Date` objects.
- Locations were stored as text strings. See the subsection below on spatial data harmonisation for more details.
- Any additional information was extracted from the property descriptions and saved as a character vector. This very often contained valuable insights not captured in the primary fields.

Data from 2012 up until the present (December 2024) were managed to be collected using this method, averaging around 1,500 listings per year. 
While highly efficient, this process relied heavily on the consistency of the site structure of the source webpages. Changes to website layouts or closures over time required significant updates to the scraping scripts. 
To overcome these issues, alternative approaches, such as using Large Language Models (LLMs) were considered. 
This is explained in the next subsection.

@@

As previously mentioned, the web scraping process also captured unstructured information from the property descriptions, which often contained valuable details not captured in the primary fields. 
In this subsection, we detail the data cleaning process using a pre-trained large language model (LLM) to extract structured information from the unstructured text. 
<!-- The LLM used was the cloud-based OpenAI o1-mini reasoning model[^1], accessed using the `{ellmer}` R package [@wickham2025ellmer] via an API. -->
The LLM used was a DeepSeek R1 distilled Qwen 14B [@deepseek2025deepseekr1;@qwen2025qwen25], accessed using the `{ellmer}` R package [@wickham2025ellmer] via the Ollama[^2] API, a local interface platform to the LLM.

@@

The primary goal was to extract the house characteristics of interest, specifically variables 7 to 15 as per @tbl-codebook, from the unstructured verbose descriptions scraped from property listing websites. 
Each description was processed with a carefully designed prompt (@fig-llm-prompt) to ensure consistent output. 
This prompt instructed the model to return only the required information in a semicolon-separated format, while handling edge cases such as missing descriptions, non-residential (commercial) properties, or rental property advertisements.

@@

The verbose descriptions were fed into the model one at a time using a loop, with the LLM extracting and returning the relevant details. 
Note that this loop was not parallelised, as the LLM already utilises multiple cores, and further parallelisation would not yield significant efficiency gains due to resource constraints.
The cleaned results were then parsed and stored in a data frame, which was then subjected to manual data-type validation to ensure conformity with the existing data set. 
It takes, on average, 81.7 seconds to process a single description using a Mac Pro 3.2GHz 16 core Intel Xeon W.
We processed `{r} scales::comma(sum(hsp$method %in% c("web-scrape-llm")))` descriptions from 2020 to 2025 using this method, with a run time of approximately 20 hours in total (spread over three machines).

@@

Our experiments show that the LLM data extraction process with the deepseek-r1:14b model achieved an overall accuracy rate of 96.9% (see [Appendix](#sec-apx-llm) for complete results).
Errors mostly stemmed from incorrect `status` classification, most likely due to the vagueness of the advertisement listing.
As for the rest of the variables, obvious errors were flagged and corrected manually during our data validation process (described in the subsection following the next one).
Overall, the LLM was found to be a valuable tool for extracting structured information from unstructured text, significantly reducing the time and effort required for data extraction. 
<!-- Users may wish to exclude these records from their analysis if they are concerned about the accuracy of the extracted data. -->

## Spatial Data Harmonisation

Whether the data was collected manually, through web scraping, or cleaned using the LLM, the spatial information extracted was often inconsistent in terms of naming conventions and granularity. 
To address this, a spatial data harmonisation process was conducted to standardise the names of the kampongs (villages) in the data set to the format used by Department of Economic Planning and Statistics (DEPS), Ministry of Finance and Economy, Brunei Darussalam as per the most recent census [@deps2022population]. 
This is the same format used by the R package `{bruneimap}` [@jamil2024bruneimap].
The CSV file `bn_kpg_level_data.csv` obtained from this package was used as a reference to standardise the kampong names in the data set, which conveniently also included the mukim and district names for each kampong.

@@

## Data Validation

@@

1. **Outlier detection.** Summary statistics analyses to identify and flag anomalous values. For instance, a built-up area recorded as 0.1 square feet or beds and baths exceeding 20 is suspect, and highly likely points towards human error or model hallucinations. Such anomalies were flagged and manually reviewed for correction.

2. **Internal consistency checks.** This made use of substantive knowledge about Brunei's housing market [@jamil2025spatiotemporal]. An example is using the price per square foot indicator, whose value typically falls within a known range. Therefore any deviations were scrutinised to ensure that variables such as price and floor area were correctly recorded. This was similarly applied to other variables such as plot area, beds, and baths.

3.	**Duplicate records detection.** We also performed duplicate record detection to identify any repeated entries that might have arisen from overlapping data sources or transcription errors. Any duplicates that were identified were carefully reviewed and removed to ensure that each property listing was uniquely represented in the span of one calendar month.

@@

The authors gratefully acknowledge the contributions of Atikah Farhain Yahya, Nurulhanisah Abdul Manan, and Nina Zuhairi towards the collection and processing of the data contained within.
We also thank the Brunei Darussalam Central Bank (BDCB) for the engaging discussions and support, which were instrumental in the inception of this project.

@@

The results show that reasoning models outperform the non-reasoning models, with OpenAI's o1-mini model performing the best (99.2% accuracy) followed by the locally run deepseek-r1:14b Ollama model (96.9% accuracy).
Most inaccuracies are seen in the `status` variable, where the models had a difficult time parsing the correct build status from the vague verbose advertisement.
However, key variables such as `price`, `type`, `plot_area`, `floor_area`, `beds`, and `baths` are generally reliable (>90% accuracy).

@@

The run time statistics for each model is shown in the table below.
The computer used Apple MacBook Air M2 Silicon on Chip (SoC) and System in Chip (SiP) processors with 16GB RAM.
Running OpenAI's models are much faster than running models locally, though a paid API key is required.
Faster local running times are, of course, a function of compute power.

@@

<!-- Unit: seconds -->
 <!--            expr       min         lq       mean     median         uq       max neval -->
 <!--        llama3.2  1.077457   1.168078   6.396146   9.595035   9.881765  10.16348    10 -->
 <!--         mistral  2.661627   2.854717  10.437078   2.907932  21.778689  22.21225    10 -->
 <!--            phi4  3.722881   3.925309  14.108101   4.267815  36.986746  38.01499    10 -->
 <!--  deepseek-r1:8b 36.829881  46.254848  65.973720  54.605889  85.101693 112.77667    10 -->
 <!-- deepseek-r1:14b 59.400884   74.62474 117.666112  81.71249   99.57201  283.8539  10 -->
 <!-- deepseek-r1:14b 40.8939 67.19222 127.0108 93.09138 151.2027 541.1301   100 -->
<!--           gpt-4o  1.279267   1.502509   2.457973   1.733497   2.395108   5.379483     5 a   -->
<!--          o1-mini  5.968378   6.551901   7.945925   7.506522   9.759592   9.943231     5 a   -->

>>>>

---
title: 'From Archives to AI: Residential Property Data Across Three Decades in Brunei Darussalam'
authors:
  - name: Haziq Jamil
    orcid: 0000-0003-3298-1010
    email: haziq.jamil@ubd.edu.bn
    url: 'https://haziqj.ml'
    corresponding: true
    affiliations:
      - name: Universiti Brunei Darussalam
        id: ubd
        department: Mathematical Sciences, Faculty of Science
        address: Universiti Brunei Darussalam, Jalan Tungku Link
        city: Bandar Seri Begawan
        country: Brunei
        postal-code: BE 1410
  - name: Amira Barizah Noorosmawie
    affiliations:
      - ref: ubd        
  - name: Hafeezul Waezz Rabu
    affiliations:
      - ref: ubd
  - name: Lutfi Abdul Razak
    orcid: 0000-0003-1681-0895
    email: lutfi.razak@ubd.edu.bn
    affiliations:
      - name: UBD School of Business and Economics
        id: ubdsbe
        # department: Mathematical Sciences, Faculty of Science
        address: Universiti Brunei Darussalam, Jalan Tungku Link
        city: Bandar Seri Begawan
        country: Brunei
        postal-code: BE 1410         
date-modified: last-modified
abstract: |
  This article introduces the first publicly available data set for analysing the Brunei housing market, covering more than 30,000 property listings from 1993 to early 2025. The data set, curated from property advertisements in newspapers and online platforms, includes key attributes such as price, location, property type, and physical characteristics, enriched with area-level spatial information. Comprehensive and historical, it complements the Brunei Darussalam Central Bank’s Residential Property Price Index (RPPI), addressing the limitations of restricted access to raw RPPI data and its relatively short timeline since its inception in 2015. Data collection involved manual transcription from archival sources and automated web scraping using programmatic methods, supported by innovative processing with Large Language Models (LLMs) to codify unstructured text. The data set enables spatial and temporal analysis, with potential applications in economics, urban planning, and real estate research. Although listing prices serve only as a proxy for market values and may deviate from actual sale prices due to negotiation dynamics and other factors, this data set still provides a valuable resource for reproducible analysis of housing market trends and for informing policy decisions.
keywords:
  - Housing Market
  - Brunei
  - Property Listings
  - Spatial Data
  - Web Scraping
  - Large Language Models
---

@@

- **First of its kind for Brunei**. This data set is the first publicly available compilation of Brunei’s residential property data, covering `{r} scales::comma(nrow(hsp))` listings from `r min(years)` to early `r max(years)`. Previous studies on Brunei's housing have either relied on recent or non-spatial data [@ng2022characterising;@jamil2025leveraging], or have primarily employed qualitative methods due to the absence of structured data [@hassan2011cultural;@hassan2023sociocultural]. Researchers may use this data to explore temporal variations in property listings and to compare housing characteristics across different regions within Brunei.

- **Historical and spatial insights**. Spanning over three decades, this data set allows for the study of long-term housing trends in Brunei. The spatial information available enables analysis at various administrative levels, providing insights into urban development patterns, regional differences, and the evolution of housing characteristics over time. This historical depth is particularly valuable given the lack of data before the establishment of the Residential Property Price Index (RPPI) [@bdcb2021technical] in 2015.

@@

- **Opportunities for handling of missing data**. The data set includes instances of missing data, particularly in house characteristics, which present opportunities for further research on data imputation techniques. Researchers may explore advanced statistical and machine learning methods to address missing values and improve data reliability. The structured format and detailed documentation of the data set support such methodological investigations, making it a valuable testbed for developing and evaluating new approaches to handling incomplete data in real-world data sets.

@@

Finally, metadata about the property is included for transparency and informational purposes. 
The variable `agent` specifies the (anonymised) identifier of the real estate agent or agency responsible for the listing, while `source` identifies the platform or medium from which the listing was obtained, such as a newspaper, magazine, or website. 
The `method` variable details the data collection approach, which is further elaborated in the section below.

@@

The date variable in the data set refers to the date on which the property listing was obtained. 
It is important to note that this is not the date the property was sold, nor does it necessarily reflect the precise timing of any transactions related to the property. 
Users should set their expectations accordingly, as the primary purpose of the date is to capture the state of the housing market at a specific point in time.

For analysis, we recommend aggregating data by quarters, as represented by the `quarter` variable. 
This aggregation helps address potential issues like missing data (see subsection below) and provides a more stable and robust representation of market trends, making it suitable for temporal analysis of the housing market.

@@

In any data collection effort, it is not surprising to encounter missing data. 
Likewise for our data set, we faced the challenge of missing values across several property characteristics--such as plot area, floor area, beds, baths, and others.
A preliminary analysis of the missing data patterns indicates that the missingness is not completely random, with certain variables displaying dependencies on others.
Advertisers often include only the information they deem most marketable or necessary, while other details may be omitted if they are considered standard or implied. 
For instance, a listing might specify the square footage and price but leave out the number of bedrooms and bathrooms, assuming that prospective buyers can infer these details.

{{< embed notebooks/missing.qmd#fig-missing-pattern >}}

Missing values are represented by blank cells in the CSV file, and the severity of missingness is summarized in @tbl-avail. 
In total, 10.1% of the records contain missing values for all key house characteristics (i.e. plot area, floor area, beds, and baths), which, depending on the research question, may require imputation or the exclusion of these records.

# Experimental Design, Materials and Methods {#sec-method}

@@

- `price` variables were cleaned by removing non-numeric characters and converted to integers.
- `beds` and `baths` were converted to integers.
- `date` variables were formatted properly as `Date` objects.
- Locations were stored as text strings. See the subsection below on spatial data harmonisation.
- Any additional information was extracted from the property descriptions and saved as a character vector. This very often contained valuable insights not captured in the primary fields.

Data from 2012 up until the present (January 2025) were managed to be collected using this method, averaging around 1,500 listings per year. 
While highly efficient, this process relied heavily on the consistency of the site structure of the source webpages. Changes to website layouts or closures over time required significant updates to the scraping scripts. 
To overcome these issues, alternative approaches, such as using Large Language Models (LLMs) were considered. 
This is explained in the next subsection.

@@

As previously mentioned, the web scraping process also captured unstructured information from the property descriptions, which often contained valuable details not captured in the primary fields. 
In this subsection, we detail the data extraction process using a pre-trained Large Language Model (LLM) to extract structured information from the unstructured text. 
<!-- The LLM used was the cloud-based OpenAI o1-mini reasoning model[^1], accessed using the `{ellmer}` R package [@wickham2025ellmer] via an API. -->
The LLM used was the DeepSeek R1 distilled Qwen 14B [@deepseek2025deepseekr1;@qwen2025qwen25], accessed using the `{ellmer}` R package [@wickham2025ellmer] via the Ollama[^2] API, a local interface platform to the LLM.

@@

The primary goal was to extract the house characteristics of interest, specifically variables 7 to 15 as per @tbl-codebook, from the unstructured verbose descriptions scraped from property listing websites. 
Each description was processed with a carefully designed prompt (@fig-llm-prompt) to ensure consistent output. 
This prompt instructed the model to return only the required information in a semicolon-separated format, while handling edge cases such as missing descriptions or non-residential (commercial) properties.

@@

The verbose descriptions were fed into the model one at a time using a loop, with the LLM extracting and returning the relevant details. 
Note that this loop was not parallelised, as the LLM already utilises multiple cores, and further parallelisation would not yield significant efficiency gains due to resource constraints.
The cleaned results were then parsed and stored in a data frame, which was then subjected to manual data-type validation to ensure conformity with the existing data set (see the last subsection). 
It takes, on average, 81.7 seconds to process a single description using a Mac Pro 3.2GHz 16 core Intel Xeon W with 48GB of DDR4 RAM.
We processed `{r} scales::comma(sum(hsp$method %in% c("web-scrape-llm")))` descriptions from 2020 to 2025 using this method, with a run time of approximately 20 hours in total (spread over three machines).

@@

Our experiments show that the LLM data extraction process with the deepseek-r1:14b model achieved an overall accuracy rate of 96.9% (see Appendix for complete results).
Errors mostly stemmed from incorrect `status` classification, likely due to the vagueness of the advertisement listing.
As for the rest of the variables, obvious errors were flagged and corrected manually during our data validation process (described in the subsection following the next one).
Overall, the LLM was found to be a valuable tool for extracting structured information from unstructured text, significantly reducing the time and effort required for data extraction. 
<!-- Users may wish to exclude these records from their analysis if they are concerned about the accuracy of the extracted data. -->

## Spatial Data Harmonisation {#sec-spatial-harm}

Whether the data was collected manually, through web scraping, or cleaned using the LLM, the spatial information extracted was often inconsistent in terms of naming conventions and granularity. 
To address this, a spatial data harmonisation process was conducted to standardise the names of the kampongs (villages) in the data set to the format used by Department of Economic Planning and Statistics (DEPS), Ministry of Finance and Economy, Brunei Darussalam as per the most recent census [@deps2022population]. 
This is the same format used by the R package `{bruneimap}` [@jamil2024bruneimap].
The CSV file `bn_kpg_level_data.csv` obtained from this package was used as a reference to standardise the kampong names in the data set, which conveniently also includes the mukim and district names for each kampong.

@@

## Data Validation {#sec-data-validation}

@@

1. **Outlier detection.** Summary statistics analyses to identify and flag anomalous values. For instance, a built-up area recorded as 0.1 square feet or beds and baths exceeding 20 is suspect, and highly likely points towards human error or model hallucinations. Such anomalies were flagged and manually reviewed for correction.
2. **Internal consistency checks.** This made use of substantive knowledge about Brunei's housing market [@jamil2025spatiotemporal]. An example is using the price per square foot indicator, whose value typically falls within a known range. Therefore any deviations were scrutinised to ensure that variables such as price and floor area were correctly recorded. This was similarly applied to other variables such as plot area, beds, and baths.
3.	**Duplicate records detection.** We also performed duplicate record detection to identify any repeated entries that might have arisen from overlapping data sources or transcription errors. Any duplicates that were identified were carefully reviewed and removed to ensure that each property listing was uniquely represented in the span of one calendar month.

@@

The authors gratefully acknowledge the contributions of Atikah Farhain Yahya, Nurulhanisah Abdul Manan, and Nina Zuhairi towards the collection and processing of the data contained within.
We also thank the Brunei Darussalam Central Bank (BDCB) for the engaging discussions and support, which were instrumental in the initiation of this project.

@@

OpenAI's models topped the accuracy charts, with the o1-mini and gpt-40 models achieving the accuracy scores of 99.2% and 98.9% respectively.
The reasoning model deepseek-r1:14b was the best performing locally run model, scoring 96.9% accuracy.
Evidently, the smaller the model, the less accurate the extraction process (cf. llama3.2 70.8%).
Most inaccuracies were seen in the `status` variable, where the models had a difficult time parsing the correct build status from the vague verbose advertisement.
However, key variables such as `price`, `type`, `plot_area`, `floor_area`, `beds`, and `baths` are generally reliable (>90% accuracy).

@@

The run time statistics for each model is shown in the table below.
The computer used was a Mac Pro 3.2GHz 16 core Intel Xeon W with 48GB of DDR4 RAM. 
At the time of running the tests, graphics card support was not available for the LLM models, which would have significantly reduced the run time.
Certainly cloud-based models such as OpenAI's models are much faster than running models locally, though a paid API key is required.

@@
